Measuring causal influence with\\ back-to-back regression: the linear case - supplementary material

### 2019-09-25 paper/synthetic.tex

B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
features. In contrast, the Forward, Backward, CCA and PLS models lead to a
loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
model). To transform such vector into an estimated causal contribution $\hat E$,
we take the sum of square coefficients:

\begin{equation} \hat E_i = \sum_j {H^j_i}^2 \end{equation}

To estimate whether models accurately identify causal factors, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}

The AUC allows evaluating the capacity of models at detecting the causal
importance of factors when ground truth labels are available, as is the case in
this setup.

We report AUC results in Figures~\ref{fig:auc_plots}~(left) and ~\ref{fig:percondition}~(top), and compare favorably to all baselines.

\begin{figure}[ht]
  \centering
  \includegraphics[width=\linewidth]{figures/AUC_R.png}
  \vspace{-4ex}
  \caption{Synthetic experiments. Distribution (over conditions) of differences in AUC (left) and Feature Importance $\Delta R$ (right) metrics between our method and the baselines. Mass under the dashed line corresponds to B2B outperforming the baseline. \label{fig:auc_plots}}
  \vspace{2ex}
  \includegraphics[width=\linewidth]{figures/AUC_conditions}
  \includegraphics[width=\linewidth]{figures/R_conditions}
  \vspace{-4ex}
  \caption{Synthetic experiments. Average AUC (top) and Feature Importance $\Delta R$ (bottom) metrics as a function of experimental conditions. Higher is better. \label{fig:percondition}}
\end{figure}

@@

For B2B, feature importance is assessed as follows:

\begin{algorithm}[H]
    %\SetAlgoLined
    \KwIn{$X_{train} \in \mathbb{R}^{n \times d_x}$, $X_{test} \in
\mathbb{R}^{n' \times d_x}$, $Y_{train} \in \mathbb{R}^{n\times d_y}$, $Y_{test}
\in \mathbb{R}^{n'\times d_y}$, } \KwOut{estimate of prediction improvement
$\Delta{R} \in \mathbb{D}^{d_x}$.} $H, G = \text{B2B}(X_{train}, Y_{train})$\;
$R_{full} = \text{corr}(X_{test} H, Y_{test} G)$\;

    \For{$i = 1, \ldots, d_x$}{ $K = Id$\; $K[i] \leftarrow 0$\; $H_k =
\text{LinearRegression}(X_{train} K, Y_{train} G_i)$\; $R_{k} =
\text{corr}(X_{test} H_k, Y_{test} G_i)$\; $\Delta R_i = R_{full} - R_{k}$\; }
\Return{$\Delta R$} \caption{B2B feature importance.} \label{algorithm:b2b_fi}
\end{algorithm}

Feature importance can be computed similarly for the CCA, PLS and Forward baselines, and we provide similar algorithm boxes in the appendix. For the Backward Model, feature importance cannot be assessed as there is no prediction.

We show in Figures~\ref{fig:auc_plots} (right) and~\ref{fig:percondition} (bottom) that our method compares favorably to baselines.

%
% \iffalse
% For CCA and PLS, feature importance is assessed as follow:
% \begin{enumerate}
%   \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
%   \item For each feature $i$:
%   \begin{itemize}
%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
%     \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
%   \end{itemize}
% \end{enumerate}
% \fi

For the Backward Model, feature importance cannot be assessed because there is no prediction.
% We thus simply
% report the model performance $R_i = corr(YG_i, X_i)$.

@@

Figure~\ref{table:synthetic} summarizes the results for all experimental
conditions, metrics, and methods.
%
B2B is the method obtaining best test errors, lowest false positive and false
negative discovery about $E$, and overall the best AUC when it comes to
detecting causal influences.

>>>>

B2B leads to unbiased (i.e. zeros-centered) scalar coefficients for non-causal
features. In contrast, the Forward, Backward, CCA and PLS models lead to a
loading vector $H_i$ per feature $i$ (or one vector $G^i$ for the backward
model). To transform such vector into an estimated causal contribution $\hat E$,
we take the sum of square coefficients:
% \begin{equation}
  $\hat E_i = \sum_j {H^j_i}^2 $
% \end{equation}

To estimate whether models accurately identify causal factors, we compute the
area-under-the-curve (AUC) across factors $AUC(E, \hat E)$.
%\begin{equation} AUC = 1 - \sum_1^n (E_k - E_{k-1}) ( \hat{E}_k +
%\hat{E}_{k-1}) / 2 \end{equation}
The AUC allows evaluating the capacity of models at detecting the causal
importance of factors when ground truth labels are available, as is the case in
this setup.

We report AUC results in Figures~\ref{fig:auc_plots}~(left) and ~\ref{fig:percondition}~(top, in Appendix), and compare favorably to all baselines.

\begin{figure}[t]
  \centering
  \includegraphics[width=\linewidth]{figures/AUC_conditions}
  \includegraphics[width=\linewidth]{figures/R_conditions}
  \vspace{-4ex}
  \caption{Synthetic experiments. Average AUC (top) and Feature Importance $\Delta R$ (bottom) when varying experimental conditions individually. Higher is better. B2B compares favorably in all cases. \label{fig:percondition}}
\end{figure}

@@

We show in Appendix~\ref{appendix:feature_importance} pseudo-code to assess feature importance for our algorithm as well as baselines. For the Backward Model, feature importance cannot be assessed as there is no prediction.

% Feature importance can be computed similarly for the CCA, PLS and Forward baselines (see Appendix \ref{appendix:feature_importance}). For the Backward Model, feature importance cannot be assessed as there is no prediction.

We show in Figures~\ref{fig:percondition} (bottom) and \ref{fig:auc_plots} (right, in Appendix) that our method compares favorably to baselines.

%
% \iffalse
% For CCA and PLS, feature importance is assessed as follow:
% \begin{enumerate}
%   \item Fit $H$ and $G$ given $X_{train}$ and $Y_{train}$
%   \item For each feature $i$:
%   \begin{itemize}
%     \item Define $K$, and identity matrix whose row $i$ has been zeroed-out
%     \item Fit $H^i_k$ and $G_k$ given $X_{train} K$ and $Y_{train}$
%   \end{itemize}
% \end{enumerate}
% \fi
% For the Backward Model, feature importance cannot be assessed because there is no prediction.
% We thus simply
% report the model performance $R_i = corr(YG_i, X_i)$.

@@

% Figure~\ref{table:synthetic} summarizes the results for all experimental
% conditions, metrics, and methods.
% %
% B2B is the method obtaining best test errors, lowest false positive and false
% negative discovery about $E$, and overall the best AUC when it comes to
% detecting causal influences.
